---
title: "Colon_DEA_groups"
author: "Narine Fischer"
date: "2025-09-05"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
## Introduction

The following workflow uses R language [@rsoftware] to carry out a Differential Expression Analysis (DEA) using data obtained from the Genomic Data Commons Data Portal [@gdcportal]. RNA-seq expression data in samples of patients with Colon Cancer (CC) was analyzed against healthy samples.

## Needed libraries

The libraries used come from both CRAN [@cran] and Bioconductor [@bioconductor]. 

```{r libraries}
library(TCGAbiolinks)
library(edgeR)
library(geneplotter)
library(limma)
library(sva)
library(dplyr)
library(ggplot2)
library(tidyr)
library(rlist)
library(openxlsx)
library(SummarizedExperiment)
library(AnnotationDbi)
library(org.Hs.eg.db)
```

## Data download

As previously stated, the data will be obtained from the Genomic Data Commons data portal [@gdcportal], a repository with data from projects studying cancer. In this case, the data will be downloaded from the project TCGA-COAD, with data specific to colon cancer. Therefore, The results published here are in whole or part based upon data generated by The Cancer Genome Atlas managed by the NCI and NHGRI. Information about TCGA can be found at http://cancergenome.nih.gov .The RNA-seq data obtained during this project will be the one selected. For this step, the package `TCGAbiolinks` will be used [@TCGAbiolinks].
```{r data_download}
# Generate query to obtain the RNA-Seq data from the TCGA-GBM project
query_rnaseq <- GDCquery(
  project = "TCGA-COAD",
  data.category = "Transcriptome Profiling",
  data.type = "Gene Expression Quantification",
  experimental.strategy = "RNA-Seq",
  access = "open"
)

# Download the RNA-Seq data
GDCdownload(query_rnaseq, method = "client")

# Prepare RNA-Seq data for analysis, creating a SummarizedExperiment object
rnaseq_data <- GDCprepare(query_rnaseq)
```

# Quality assessment and processing of data

## Observation of possible confounding variables

As the intent of this workflow is to compare primary tumor samples against healthy samples, the `rnaseq_data` will be filtered to only obtain these two types of samples. Moreover a DGELISt object will be created for further analyses

```{r filter_rnsaseq_data}
# Filter for healthy and primary tumor samples
se<- rnaseq_data[, colData(rnaseq_data)$definition %in% c("Solid Tissue Normal", "Primary solid Tumor")]

# Create a DGEList object for the analysis
dge <- DGEList(counts=assay(se), genes=rowData(se),
                 samples=colData(se))
```

The sequencing depth per sample will be mapped and colored according to a variable that is believed that might cause a confounding effect. The first variable to be observed is `gender`. The aim is to see if there is any difference in sequencing depth due to gender, this should be done with any variable that wants to be investigated. A graph with the ordered sequencing depths and colored by `gender` is created.

```{r quality_assessment_gender}
# Quality assessment: depth sequencing - gender
ord <- order(dge$sample$lib.size)
se$gender<-factor(se$gender)
barplot(dge$sample$lib.size[ord]/1e6, las=1, ylab="Millions of reads",
        xlab="Samples", col=c("red", "blue", "white")[se$gender[ord]])
legend("topleft", c("female", "male", "not reported"), fill=c("red", "blue", "white"), inset=0.01)
```

There are some considerable differences in sequencing depth per sample, ranging from less than 20 to more than 80 million reads; therefore, normalization within samples is needed. In addition, it doesn't appear to be differences in sequencing depth due to gender.

The next variable to be assessed is `race`.

```{r quality_assessment_race}
# Quality assessment: depth sequencing - race. A graph with the ordered sequencing depths and colored by race is created.
ord <- order(dge$sample$lib.size)
se$race<-factor(se$race)
barplot(dge$sample$lib.size[ord]/1e6, las=1, ylab="Millions of reads",
        xlab="Samples", col=c("green", "yellow", "#ED3EF7","white", "#FF885B")[se$race[ord]])
legend("topleft", c("american indian or alaska native","asian", "black or african american", "not reported", "white"), fill=c("green", "yellow", "#ED3EF7", "white", "#FF885B"), inset=0.01)
```

Whereas gender was more homogeneous, there are clear differences between races. "Not reported" samples are mainly in the lower side of sequencing depth, while the rest of samples accumulate on the top of the sequencing depth. Race will be taken as a covariate.

It will also be seen if there is any distinction in sequencing depth depending on the target variable, `definition`.
```{r quality_assessment_definition}
# Quality assessment: depth sequencing - definition. A graph with the ordered sequencing depths and colored by definition is created
se$definition<-factor(se$definition)
barplot(dge$sample$lib.size[ord]/1e6, las=1, ylab="Millions of reads",
        xlab="Samples", col=c("brown", "lightpink")[se$definition[ord]])
legend("topleft", c("Primary solid Tumor", "Solid Tissue Normal"), fill=c("brown", "lightpink"), inset=0.01)
```

It can be seen that Solid tissue normal is mainly present in the higher region of sequencing depth, confirming that normalization is necessary.

## Normalization I: Within-sample

In order to adjust for the differences in sequencing depth, normalization within samples will be carried out. For this, counts per million (CPM) will be calculated. Moreover, the logarithm with base 2 will be carried out to reduce variability and make the data more suitable for statistical analysis.

```{r within_sample_normalization}
# Calculate the log2 of the CPM and add it to the SummarizedExperiment object 
assays(se)$logCPM <- cpm(dge, log=TRUE, prior.count=0.25)
```

The distribution of the expression of the `logCPM` values is plotted to identify if there are any samples that deviates from the rest.

```{r QA_sample_distribution}
# Representation of the distribution of logCPM
par(mfrow=c(1, 2), mar=c(4, 5, 4, 4))
multidensity(as.list(as.data.frame(assays(se)$logCPM)),
               xlab="log2 CPM", legend=NULL,
               main="", cex.axis=1.2, cex.lab=1.5, las=1)
boxplot(assays(se)$logCPM, col="gray", xlab="Samples",
               ylab=expression(log[2] * "CPM"), cex.axis=1.2, cex.lab=1.5, las=1)
```

It appears that there are some unsuitable samples. These samples will be removed.

```{r detecting_samples_to_remove}
num_matrix <- as.matrix(assay(se, "logCPM"))

density_0<- function(x){
  d <-density(x, n=4096)
  # infer density at -1
  unname(approx(d$x, d$y, xout=-1, rule=2)$y)
}

densities<- vapply(seq_len(ncol(num_matrix)), function(j) density_0(num_matrix[,j]), numeric(1))

names(densities) <- colnames(num_matrix)

samples_to_erase <- names(densities[densities >0.05])

# replot to mkae sure the correct samples have been identified

replot_density <- function(full_matrix, highlight){
  # add smoothness to curves
  style<-par(lend=2)
  on.exit(par(style))
  
  # create a blank graph 
  plot (0,0,type="n", xlab="log2 CPM", ylab="Density", xlim=range(full_matrix), ylim=c(0,0.5))
  
  #fill the graph
  for (s in colnames(full_matrix)){
    densities<-density(full_matrix[,s], n=2048)
    #choose colour
    colour<- if (s %in% highlight) "red" else "grey"
    lines(densities$x, densities$y, lwd=if(s %in% highlight) 2.5 else 1, col=colour)
  }
  legend("topright", lwd=c(2.5,1), col=c("red", "grey"), legend= c("eliminate", "save"), bty="n")
}

replot_density(num_matrix, samples_to_erase)
```

It can be seen that the samples detected indeed are the samples to remove, the removal is carried out. In addition, the samples were searched in the sample datatable to see if they were cancer samples, or healthy samples. All of them are cancer samples, making the total of samples after removal: 79 cancer samples and 197 healthy samples.

```{r remove_bad_samples}
se <- se[, !colnames(se) %in% samples_to_erase]
dge <- DGEList(counts=assay(se), genes=rowData(se),
                 samples=colData(se))
```

Finally, we replot the multidensity plot to make sure no other unsuitable samples remain.

```{r make_sure}
# Representation of the distribution of logCPM
par(mfrow=c(1, 2), mar=c(4, 5, 4, 4))
multidensity(as.list(as.data.frame(assays(se)$logCPM)),
               xlab="log2 CPM", legend=NULL,
               main="", cex.axis=1.2, cex.lab=1.5, las=1)
boxplot(assays(se)$logCPM, col="gray", xlab="Samples",
               ylab=expression(log[2] * "CPM"), cex.axis=1.2, cex.lab=1.5, las=1)
```

It can be seen that there are still some unsuitable samples.  A second round of elimination is done. 

```{r detecting_samples_to_remove}
num_matrix <- as.matrix(assay(se, "logCPM"))

density_0<- function(x){
  d <-density(x, n=4096)
  # infer density at -3
  unname(approx(d$x, d$y, xout=-3, rule=2)$y)
}

densities<- vapply(seq_len(ncol(num_matrix)), function(j) density_0(num_matrix[,j]), numeric(1))

names(densities) <- colnames(num_matrix)

samples_to_erase <- names(sort(densities, decreasing=FALSE))[1:3]

# replot to mkae sure the correct samples have been identified

replot_density <- function(full_matrix, highlight){
  # add smoothness to curves
  style<-par(lend=2)
  on.exit(par(style))
  
  # create a blank graph 
  plot (0,0,type="n", xlab="log2 CPM", ylab="Density", xlim=range(full_matrix), ylim=c(0,0.5))
  
  #fill the graph
  for (s in colnames(full_matrix)){
    densities<-density(full_matrix[,s], n=2048)
    #choose colour
    colour<- if (s %in% highlight) "red" else "grey"
    lines(densities$x, densities$y, lwd=if(s %in% highlight) 2.5 else 1, col=colour)
  }
  legend("topright", lwd=c(2.5,1), col=c("red", "grey"), legend= c("eliminate", "save"), bty="n")
}

replot_density(num_matrix, samples_to_erase)
```

```{r remove_bad_samples}
se <- se[, !colnames(se) %in% samples_to_erase]
dge <- DGEList(counts=assay(se), genes=rowData(se),
                 samples=colData(se))
```

```{r detecting_samples_to_remove}
num_matrix <- as.matrix(assay(se, "logCPM"))

density_0<- function(x){
  d <-density(x, n=4096)
  # infer density at -1
  unname(approx(d$x, d$y, xout=-2.5, rule=2)$y)
}

densities<- vapply(seq_len(ncol(num_matrix)), function(j) density_0(num_matrix[,j]), numeric(1))

names(densities) <- colnames(num_matrix)

samples_to_erase <- names(densities[densities >0.06])

# replot to mkae sure the correct samples have been identified

replot_density <- function(full_matrix, highlight){
  # add smoothness to curves
  style<-par(lend=2)
  on.exit(par(style))
  
  # create a blank graph 
  plot (0,0,type="n", xlab="log2 CPM", ylab="Density", xlim=range(full_matrix), ylim=c(0,0.5))
  
  #fill the graph
  for (s in colnames(full_matrix)){
    densities<-density(full_matrix[,s], n=2048)
    #choose colour
    colour<- if (s %in% highlight) "red" else "grey"
    lines(densities$x, densities$y, lwd=if(s %in% highlight) 2.5 else 1, col=colour)
  }
  legend("topright", lwd=c(2.5,1), col=c("red", "grey"), legend= c("eliminate", "save"), bty="n")
}

replot_density(num_matrix, samples_to_erase)
```

```{r remove_bad_samples}
se <- se[, !colnames(se) %in% samples_to_erase]
dge <- DGEList(counts=assay(se), genes=rowData(se),
                 samples=colData(se))
```

```{r make_sure}
# Representation of the distribution of logCPM
par(mfrow=c(1, 2), mar=c(4, 5, 4, 4))
multidensity(as.list(as.data.frame(assays(se)$logCPM)),
               xlab="log2 CPM", legend=NULL,
               main="", cex.axis=1.2, cex.lab=1.5, las=1)
boxplot(assays(se)$logCPM, col="gray", xlab="Samples",
               ylab=expression(log[2] * "CPM"), cex.axis=1.2, cex.lab=1.5, las=1)
```

Although it is much better, the last two samples that are deemed unsuitable will be eliminated.

```{r detecting_samples_to_remove}
num_matrix <- as.matrix(assay(se, "logCPM"))

density_0<- function(x){
  d <-density(x, n=4096)
  # infer density at -3
  unname(approx(d$x, d$y, xout=-2.5, rule=2)$y)
}

densities<- vapply(seq_len(ncol(num_matrix)), function(j) density_0(num_matrix[,j]), numeric(1))

names(densities) <- colnames(num_matrix)

samples_to_erase <- names(sort(densities, decreasing=FALSE))[1:2]

# replot to mkae sure the correct samples have been identified

replot_density <- function(full_matrix, highlight){
  # add smoothness to curves
  style<-par(lend=2)
  on.exit(par(style))
  
  # create a blank graph 
  plot (0,0,type="n", xlab="log2 CPM", ylab="Density", xlim=range(full_matrix), ylim=c(0,0.5))
  
  #fill the graph
  for (s in colnames(full_matrix)){
    densities<-density(full_matrix[,s], n=2048)
    #choose colour
    colour<- if (s %in% highlight) "red" else "grey"
    lines(densities$x, densities$y, lwd=if(s %in% highlight) 2.5 else 1, col=colour)
  }
  legend("topright", lwd=c(2.5,1), col=c("red", "grey"), legend= c("eliminate", "save"), bty="n")
}

replot_density(num_matrix, samples_to_erase)
```

```{r remove_bad_samples}
se <- se[, !colnames(se) %in% samples_to_erase]
dge <- DGEList(counts=assay(se), genes=rowData(se),
                 samples=colData(se))
```

```{r make_sure}
# Representation of the distribution of logCPM
par(mfrow=c(1, 2), mar=c(4, 5, 4, 4))
multidensity(as.list(as.data.frame(assays(se)$logCPM)),
               xlab="log2 CPM", legend=NULL,
               main="", cex.axis=1.2, cex.lab=1.5, las=1)
boxplot(assays(se)$logCPM, col="gray", xlab="Samples",
               ylab=expression(log[2] * "CPM"), cex.axis=1.2, cex.lab=1.5, las=1)
```

It can be confirmed that no more unsuitable samples remain. Therefore, the next step can begin. In the end, the samples saved are:

```{r table_se_definition}
table(factor(se$definition))
```

Before, there were 481 tumor samples, and 41 healthy samples. Now, 452 tumor samples remain, and 41 healthy samples remain. Therefore, all samples erased were tumor samples. 

# Example test

Due to the high difference between the number of tumor samples and healthy samples, for the rest of the data processing and analyses, it has been decided to perform multiple comparisons; in each one, 41 random samples of primary tumor will be compared against the 41 healthy samples. New objects will be created to store the 41 samples classified as `Solid Tissue Normal`, and 41 random samples classified as `Primary solid Tumor`. These new objects will be the ones used from this point forward. It has been decided to create them at this point because the next filtering and normalization steps are dependent on the samples selected.

For the DEA, 1000 groups of 41 Primary solid Tumor samples will be created and 1000 DEAs will be carried out to obtain the differentially expressed genes. But first, an example with one group will be executed to determine possible factors that can affect the analysis.

First, the objects containing 5 tumoral samples and 5 healthy samples are created.
```{r creating_group}
# Obtain the sample metadata
sample_metadata <- colData(se)

################### Obtaining the healthy samples
# Get all the healthy samples IDs
healthy_samples <- rownames(sample_metadata[sample_metadata$definition == "Solid Tissue Normal", ])

################## Obtaining the Primary tumor samples
# Select 41 random sample IDs from "Primary solid Tumor"
set.seed(123)  
primary_tumor_samples <- rownames(sample_metadata[sample_metadata$definition == "Primary solid Tumor", ])
random_primary_tumor_samples <- sample(primary_tumor_samples, 41)

################## Creating the SummarizedExperiment object to be used
# Combine healthy samples IDs and the randomly selected primary tumor sample IDs
selected_samples <- c(healthy_samples, random_primary_tumor_samples)

# Subset the SummarizedExperiment object with the selected samples
se_subset <- se[, colnames(se) %in% selected_samples]

# Make sure that the new object only has 10 samples (10 rows)
dim(colData(se_subset))

################## Creating the DGEList object to be used
# Subset the previous DGEList object
dge_subset <- dge[, colnames(dge) %in% selected_samples]

# Make sure that there are 10 samples (10 columns)
dim(dge_subset)
```

The dimensions of the new SummarizedExperiment and DGEList objects are printed to make sure they have been created accordingly, as 82 samples in total have been selected, the metadata for the samples in the SummarizedExperiment should have 82 rows, and the DGEList object should have 82 columns. It can be seen that this is indeed what happens, therefore, the objects have been created correctly.

## Filtering of lowly-expressed genes

The next step will be filtering out the lowly-expressed genes, as they can introduce artifacts in the analysis. First, a representation of the gene expression is done.
```{r QA_gene_distribution}
# Calculate the average level of expression of each gene
avgexp <- rowMeans(assays(se_subset)$logCPM)

# Plot the average levels of expression 
hist(avgexp, xlab=expression(log[2] * "CPM"), main="", las=1, col="gray")
abline(v=0, col="red", lwd=2)
```

It can be seen that most of the genes are lowly expressed. To continue with the analysis, the lowly-expressed genes will be removed. In this case, the cutoff has been calculated according to Chen et al. (2016), 10 reads per million in the sample with the lowest depth. Genes kept must have this minimum level of expression in at least as many samples as the smallest group of comparison (healthy samples, 41 samples)

```{r LE_genes_removal}
# Create mask using 1 log2CPM as cutoff
cpmcutoff <- round(10/min(dge$samples$lib.size/1e6),digits=1)
nsamplescutoff <- min(table(se$definition))

mask <- rowSums(cpm(dge) > cpmcutoff) >= nsamplescutoff

# Create new objects with the filtered genes 
se_subset.filt <- se_subset[mask, ]
dge_subset.filt <- dge_subset[mask, ]
dim(se_subset.filt)
```

A total of 16988 genes that meet the cutoff are kept.

Plotting again the gene expression plot, coloring the 16988 genes in red, the correct selection is confirmed
```{r filtered_genes_representation}
par(mar=c(4, 5, 1, 1))
h <- hist(avgexp, xlab=expression("Expression level (" * log[2] * "CPM)"),
          main="", las=1, col="grey", cex.axis=1.2, cex.lab=1.5)
x <- cut(rowMeans(assays(se_subset.filt)$logCPM), breaks=h$breaks)
lines(h$mids, table(x), type="h", lwd=10, lend=1, col="red")
legend("topright", c("All genes", "Filtered genes"), fill=c("grey", "red"))
```

## Normalization II: between-sample

The next step will be normalizing between samples. The TMM method will be used.
```{r normalization_bet_sample}
dge_subset.filt <- calcNormFactors(dge_subset.filt, method = 'TMM')
```

The raw data is substituted by the normalized data
```{r raw_to_normalized}
# Replace the raw log2 CPM units in the corresponding assay element of the `SummarizedExperiment` object, by the normalized ones.
assays(se_subset.filt)$logCPM <- cpm(dge_subset.filt, log=TRUE,
                              normalized.lib.sizes=TRUE)
```

A boxplot is created to see if there are any significant differences in expression between samples to see if there is any outlier between them
```{r repeat_boxplot}
boxplot(assays(se_subset.filt)$logCPM, col="gray", xlab="Samples",
        ylab=expression(log[2] * "CPM"), las=1)
```

No major difference between the samples are observed.

## QA: MA-plots

To compare, first the individual MA-plots prior to filtering will be plotted. 
```{r MA_plot_before_filter}
# Print them 4 at a time
par(mfrow=c(2, 2), mar=c(4, 5, 1, 1))

# MA-plot generation
for (i in 1:ncol(se_subset)) {
  A <- rowMeans(assays(se_subset)$logCPM)
  M <- assays(se_subset)$logCPM[, i] - A
  smoothScatter(A, M, main=colnames(se_subset)[i], las=1)
  abline(h=0, col="blue", lwd=2)
  lo <- lowess(M ~ A)
  lines(lo$x, lo$y, col="red", lwd=2)
}
```

Then, the individual MA-plots after filtering and normalization against the average values will be plotted 
```{r MA_plot_after_filter}
# Print 4 graphs at a time
par(mfrow=c(2, 2), mar=c(4, 5, 1, 1))

# MA-plot generation
for (i in 1:ncol(se_subset.filt)) {
  A <- rowMeans(assays(se_subset.filt)$logCPM)
  M <- assays(se_subset.filt)$logCPM[, i] - A
  smoothScatter(A, M, main=colnames(se_subset.filt)[i], las=1)
  abline(h=0, col="blue", lwd=2)
  lo <- lowess(M ~ A)
  lines(lo$x, lo$y, col="red", lwd=2)
}
```

It can be seen that there are some small differences between the individual sample expression trends (red line) and the average expression across all the samples (blue line). In addition, as lowly-expressed genes have been filtered out, the differences observed in lowly-expressed genes in the "before filtering" graphs has been reduced, focus on the genes with higher and more reliable levels of expression. 

Next, a smear plot will be done to compare `Primary solid Tumor` and `Solid Tissue Normal`as groups. As before, a graph before filtering and another after filtering will be carried out.
```{r comparing_by_groups}
# Print the graphs together
par(mfrow=c(2, 1), mar=c(4, 5, 1, 1))

# Assign the type of samples to the DGEList objects
se_subset$definition <- relevel(factor(se_subset$definition), ref="Solid Tissue Normal")
se_subset.filt$definition <- relevel(factor(se_subset.filt$definition), ref="Solid Tissue Normal")
dge_subset$samples$group <- se_subset$definition
dge_subset.filt$samples$group <- se_subset$definition

# Smear plot before filtering
plotSmear(dge_subset, lowess=TRUE, las=1, cex.lab=1.5, cex.axis=1.2)
abline(h=0, col="blue", lwd=2)

# Smear plot after filtering
plotSmear(dge_subset.filt, lowess=TRUE, las=1, cex.lab=1.5, cex.axis=1.2)
abline(h=0, col="blue", lwd=2)

```
In this case, the blue line represents equal expression of the genes between the two types of samples, while the red line portrays the differences in expression between the groups. Therefore, deviations of the red line from the blue line signal that there are genes that may be differentially expressed . Additionally, as this is only an example of one group, it is possible that another group with other samples might give more or less deviation.

## Batch identification and experimental design

In order to detect possible batches that can interfere in the analysis, a hierarchical  clustering is carried out.
```{r batch_identification_hierarchical clustering,  fig.height=5, fig.width=8, dpi=100,}
# Calculate log CPM values with a prior.count of 3 to moderate extreme fold-changes produced by low counts
logCPM <- cpm(dge_subset.filt, log=TRUE, prior.count=3)

# Calculate the distance matrix using Spearman correlation
d <- as.dist(1-cor(logCPM, method="spearman"))

# Perform hierarchical clustering
sampleClustering <- hclust(d)

# To observe gender as a possible batch effect, an object will be created for later use to color each sample's gender
batch <- as.integer(se_subset.filt$gender)
names(batch) <- colnames(se_subset.filt)

# Create a dendrogram object with the hierarchical clustering 
sampleDendrogram <- as.dendrogram(sampleClustering, hang=0.1)

# Create the labels that will be seen in the dendrogram
outcome <- se_subset.filt$shortLetterCode
names(outcome) <- colnames(se_subset.filt)

# Modify dendrogram labels and color
sampleDendrogram <- dendrapply(sampleDendrogram,
                               function(x, batch, labels) {
                                 if (is.leaf(x)) {
                                   attr(x, "nodePar") <- list(lab.col=as.vector(batch[attr(x, "label")]))
                                   attr(x, "label") <- as.vector(labels[attr(x, "label")])
                                 }
                                 x
                               }, batch, outcome)

# Plot the dendrogram
plot(sampleDendrogram, main="Hierarchical clustering of samples",
     cex=0.7)
legend("topright", levels(se_subset.filt$gender),
       fill=seq_len(nlevels(se_subset.filt$gender)))
```

```{r batch_identification_hierarchical clustering,  fig.height=5, fig.width=8, dpi=100,}
# Calculate log CPM values with a prior.count of 3 to moderate extreme fold-changes produced by low counts
logCPM <- cpm(dge_subset.filt, log=TRUE, prior.count=3)

# Calculate the distance matrix using Spearman correlation
d <- as.dist(1-cor(logCPM, method="spearman"))

# Perform hierarchical clustering
sampleClustering <- hclust(d)

# To observe gender as a possible batch effect, an object will be created for later use to color each sample's gender
batch <- as.integer(se_subset.filt$race)
names(batch) <- colnames(se_subset.filt)

# Create a dendrogram object with the hierarchical clustering 
sampleDendrogram <- as.dendrogram(sampleClustering, hang=0.1)

# Create the labels that will be seen in the dendrogram
outcome <- se_subset.filt$shortLetterCode
names(outcome) <- colnames(se_subset.filt)

# Modify dendrogram labels and color
sampleDendrogram <- dendrapply(sampleDendrogram,
                               function(x, batch, labels) {
                                 if (is.leaf(x)) {
                                   attr(x, "nodePar") <- list(lab.col=as.vector(batch[attr(x, "label")]))
                                   attr(x, "label") <- as.vector(labels[attr(x, "label")])
                                 }
                                 x
                               }, batch, outcome)

# Plot the dendrogram
plot(sampleDendrogram, main="Hierarchical clustering of samples",
     cex=0.7)
legend("topright", levels(se_subset.filt$race),
       fill=seq_len(nlevels(se_subset.filt$race)))
```

Another way to determine batches is through a Multidimensional Scaling (MDS) plot.
```{r batch_identificatio_nmps_graph}

batch <- as.integer(se_subset.filt$gender)
names(batch) <- colnames(se_subset.filt)

# Creating the labels of the samples
outcome <- se_subset.filt$shortLetterCode
names(outcome) <- colnames(se_subset.filt)

# Creating the MDS plot
plotMDS(dge_subset.filt, labels=outcome, col=batch)
legend("topright", levels(se_subset.filt$gender),
       fill=seq_len(nlevels(se_subset.filt$gender)), inset=0.05)
```

```{r batch_identificatio_nmps_graph}

batch <- as.integer(se_subset.filt$race)
names(batch) <- colnames(se_subset.filt)

# Creating the labels of the samples
outcome <- se_subset.filt$shortLetterCode
names(outcome) <- colnames(se_subset.filt)

# Creating the MDS plot
plotMDS(dge_subset.filt, labels=outcome, col=batch)
legend("top", levels(se_subset.filt$race),
       fill=seq_len(nlevels(se_subset.filt$race)), inset=0.05)
```

## Differential expression analysis

In order to start the DEA; first the levels of the target variable are modified to make the healthy samples as references, this way, all results will be of the tumor samples compared to the healthy samples. Moreover, the model for the analysis is created, as stated before, no confounding effect will be taken into account, as the healthy samples lack enough data to be able to do it.
```{r define_the_reference}
# Define the reference
se_subset.filt$definition <- relevel(se_subset.filt$definition, ref="Solid Tissue Normal")

# Creating the model for the analysis
mod <- model.matrix(~ definition + race, data=colData(se_subset.filt))
head(mod, 15)
```

Although some possible confounding effects were tested for, there might be others that are being missed, hidden batch effects. To avoid this, a Surrogate Variable Analysis (SVA) will be carried out.
```{r SVA}
# Creating a model to compare the previously created against
mod0 <- model.matrix(~ race, colData(se_subset.filt))

# Carry out the SVA analysis
sv <- sva(assays(se_subset.filt)$logCPM, mod=mod, mod0=mod0)
```

Thanks to the results fo the analysis, 3 significant surrogate variables are obtained.

In order to see their significance, a Singular Value Decomposition (SVD) is done to see the amount of contribution to the variance by each surrogate variable
```{r SVA_SVD}
# Perform SVD on the logCPM values
svd_result <- svd(scale(assays(se_subset.filt)$logCPM))
eigenvalues <- svd_result$d^2

# Plot eigenvalues
plot(eigenvalues, type = "b", xlab = "Component", ylab = "Eigenvalue", main = "SVD Eigenvalues")

```

The model is modified to take the surrogate variables into account. Then the DEA is carried out with `voomLmFit`from the package `EdgeR` [@edgeR], and with `eBayes`from the package `limma` [@limma]
```{r t-statistic}
# Addition of the SVA surregate variables
mod <- cbind(mod, sv$sv)

# Carry out the Differential Expression Analysis
fit <- voomLmFit(dge_subset.filt, mod, block= se_subset$patient)
fit <- eBayes(fit)
```


The results of the DEA can be found in the object `fit`.

In `fit`, the genes do not appear as their symbols, to make it more understandable, the Ensembl code [@Ensembl2024] will be changed by the genes' symbols. 
```{r gene_data}
# 1) Get Ensembl gene IDs (strip version if present)
ens <- rowData(se_subset.filt)$gene_id
ens <- sub("\\.\\d+$", "", ens)  # drop ".12" version suffixes

# 2) Map to SYMBOL, ENTREZID, chromosome, and a readable name
symbol <- mapIds(org.Hs.eg.db, keys = ens, keytype = "ENSEMBL", column = "SYMBOL",   multiVals = "first")
entrez <- mapIds(org.Hs.eg.db, keys = ens, keytype = "ENSEMBL", column = "ENTREZID", multiVals = "first")
chr    <- mapIds(org.Hs.eg.db, keys = ens, keytype = "ENSEMBL", column = "CHR",      multiVals = "first")
gname  <- mapIds(org.Hs.eg.db, keys = ens, keytype = "ENSEMBL", column = "GENENAME", multiVals = "first")

# 3) Build the data.frame for limma
genesmd <- data.frame(
  chr         = unname(chr),
  symbol      = unname(symbol),                     # or rowData(se.filt)$Gene_name if you prefer
  description = unname(gname),                      # a readable gene name/description
  EntrezID    = unname(entrez),
  stringsAsFactors = FALSE,
  row.names = rownames(se_subset.filt)
)

# 4) Attach to the limma fit
fit$genes <- genesmd
```

Before proceeding, correction for multiple testing is necessary, for it, the function `topTable()` is used. This function uses Benjamini-Hochberg FDR (Flase Discovery Rate) as method for correction.
Moreover, the amount of genes that are overexpressed, underexpressed or stays the same in healthy samples (intercept) will also be printed.
```{r multiple_test_correction}
# Multiple testing correction
tt <- topTable(fit, coef=2, n=Inf)

# Genes divided by change in expression
res <- decideTests(fit, p.value=0.1)
print(summary(res))
```

A plot with the distribution of the P.values for each gene is created.
```{r pvalues}
# Create a histogram with the P.values
h <- hist(tt$P.Value, xlab="Raw P-values", main="",
          las=1)
```

It can be seen that many genes have low p-values, signaling to differences in expression.

A volcano plot is also created to see the distribution of the differentially expressed genes, either underexpressed or overexpressed.
```{r volcano}
# Create the volcano plot
volcanoplot(fit, coef=2, highlight=10, names=fit$genes$symbol,
            main="", las = 1)
```

In addition, an MA-plot will also be created to also represent the differentially expressed genes distribution.
```{r MA_plots_after}
# Obtaining the name of the top 10 DEG
top10 <- order(fit$lods[, 2], decreasing=TRUE)[1:10]

# Filter for genes with less than a 0.1 adjusted P.value
DEgenes <- rownames(tt)[tt$adj.P.Val < 0.1]

# Createing the MA-plot
limma::plotMA(fit, coef=2, status=rownames(fit$lods) %in% DEgenes, legend=FALSE,
              main="", hl.pch=46, hl.cex=4, bg.pch=46, bg.cex=3, las =1)
text(fit$Amean[top10], fit$coef[top10, 2], fit$genes$symbol[top10], cex=0.5, pos=4)
```

To have the complete list of differentially expressed genes, a final filter is undergone, only those that after the correction have an adjusted P.value of less than 0.1 will be kept.
```{r DGE_list}
DEGtable <- tt %>% filter(adj.P.Val < 0.1) %>% arrange(P.Value)
rownames(DEGtable) <- 1:nrow(DEGtable)
```

However, as previously stated, this example was only to determine which factors to take into account for the rest of the groups, for example: using the sva data, but not use variables such as race, gender or age due to lack of information in the healthy samples. 

# Carrying out the 1000 DEAs

In order to obtain results using all the data available, multiple groups of 5 primary tumor samples will be created, and the same path will be followed (without the graphics, to optimize performance), obtaining the list of all comparisons.
```{r loop}
# Defining the objects where the results will be stored
results<- data.frame(assay=character(), symbol=character(), logFC=numeric(), adjPvalue=numeric())
samples_selected <- vector("list", 1000)

# Carry out the multiple DEAs
for (i in 1:1000){
  # Counter in order to know in which comparison it is
  print(i)
  
  # Randomly select 5 samples from "Primary solid Tumor"
  random_primary_tumor_samples <- sample(primary_tumor_samples, 41)
  samples_selected[[i]]<- random_primary_tumor_samples
  
  # Combine healthy samples and the randomly selected primary tumor samples.
  selected_samples <- c(healthy_samples, random_primary_tumor_samples)

  # Subset the SummarizedExperiment object with the selected samples
  se_subset <- se[, colnames(se) %in% selected_samples]

  # Subset the previous DGEList object
  dge_subset <- dge[, colnames(dge) %in% selected_samples]
  
  # Filtering lowly expressed genes
  mask <- rowMeans(assays(se_subset)$logCPM) > 1
  se_subset.filt <- se_subset[mask, ]
  dge_subset.filt <- dge_subset[mask, ]
  
  # Between-sample normalization
  dge_subset.filt <- calcNormFactors(dge_subset.filt, method = 'TMM')
  assays(se_subset.filt)$logCPM <- cpm(dge_subset.filt, log=TRUE,
                              normalized.lib.sizes=TRUE)
  
  # Differential expression analysis
  se_subset.filt$definition <- relevel(se_subset.filt$definition, ref="Solid Tissue Normal")
  mod <- model.matrix(~ definition, data=colData(se_subset.filt))
  mod0 <- model.matrix(~ 1, colData(se_subset.filt))
  sv <- sva(assays(se_subset.filt)$logCPM, mod=mod, mod0=mod0)
  mod <- cbind(mod, sv$sv)
  fit <- voomLmFit(dge_subset.filt, mod, block = se_subset$patient)
  fit <- eBayes(fit)
  genesmd <- data.frame(chr=as.character(seqnames(rowRanges(se_subset.filt))),
                        symbol=rowData(se_subset.filt)[, 7], description=rowData(se_subset.filt)[, 6],
                      EntrezID = rownames(se_subset.filt), stringsAsFactors=FALSE)
  fit$genes <- genesmd
  tt <- topTable(fit, coef=2, n=Inf)
  DEGtable <- tt %>% filter(adj.P.Val < 0.1) %>% arrange(P.Value)
  rownames(DEGtable) <- 1:nrow(DEGtable)
  
  # Obtaining the important information from the results
 print("adding results")
 new_entry<-cbind(i, DEGtable %>% select(symbol, logFC, adj.P.Val))
 colnames(new_entry)<-colnames(results)
 results<-rbind(results, new_entry)
}
```

# Analysing the results

After obtaining the list with all the differentially expressed genes in `results`, a count table is created to see how many times each of the wanted genes for the study appears in the results of the analyses.

```{r count_matrix_creation}
# Creating the count table
gene_counts <- table(results$symbol)

# Convert the table to a data frame
gene_count_df <- as.data.frame(gene_counts)
colnames(gene_count_df) <- c("Gene", "Count")
rownames(gene_count_df) <- gene_count_df[,1]
```

Revising the distribution of the results might give an insight of them.
```{r DEG_distribution}
# A data frame with the possible ranges of number of counts
bins <- data.frame(Count_Bin = factor(
  levels(cut(0, breaks = seq(0, 1000, by = 100), right = FALSE, include.lowest = TRUE)),
  levels = unique(cut(seq(0, 1000, by = 1), breaks = seq(0, 1000, by = 100), right = FALSE, include.lowest = TRUE))
))

# Object that will divide the genes in groups depending on the number of DEAs they ahve appeared as a DEG
gene_count_summary <- gene_count_df %>%
  mutate(Count_Bin = cut(Count, breaks = seq(0, 1000, by = 100), right = FALSE, include.lowest = TRUE)) %>%
  group_by(Count_Bin) %>%
  summarize(Num_Genes = n()) %>%
  right_join(bins, by = "Count_Bin") %>%
  replace_na(list(Num_Genes = 0))

# Plot the distribution
ggplot(gene_count_summary, aes(x = Count_Bin, y = Num_Genes)) +
  geom_bar(stat = "identity", fill = "steelblue") +
  labs(
    title = "Distribution of Gene Counts by Range",
    x = "Count Range",
    y = "Number of Genes"
  ) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

Even though multiple testing correction has been taken into account independently in each DEA, as 1000 analyses have been performed, another multiple testing correction must be applied to take into account the number of analyses carried out.  
```{r FDR}
# Creation of the object that will store the results after multiple testing correction
results_fdr <- matrix(nrow=0, ncol=4)

# Obtaining the newly adjusted P.values with FDR modifying the adjusted P.values obtained from the DEAs
for (i in 1:nrow(gene_count_df)){
  gene_pvalues<- results %>% filter(symbol == gene_count_df[i, "Gene"])
  adjusted_pvalues <- p.adjust(gene_pvalues$adjPvalue, method = "fdr", n=1000)
  copy_data <- results %>% filter(symbol==gene_count_df[i, "Gene"])
  copy_data$adjPvalue <- adjusted_pvalues
  results_fdr<- rbind(results_fdr, copy_data)
}

# Filter the adjusted P.values using a cutoff of  0.1
results_fdr <- results_fdr %>% filter(adjPvalue < 0.1)

# Count the frequency of each gene across all comparisons
gene_counts <- table(results_fdr$symbol)

# Convert the table to a data frame
gene_count_df_fdr <- as.data.frame(gene_counts)
colnames(gene_count_df_fdr) <- c("Gene", "Count")
rownames(gene_count_df_fdr) <- gene_count_df_fdr[,1]
```

Now, the object `gene_count_df_fdr` carries the number of times each gene has appeared as a DEG after eliminating the possible false positives. The distribution is plotted again.
```{r distribution_II}
# A data frame with the possible ranges of number of counts
bins <- data.frame(Count_Bin = factor(
  levels(cut(0, breaks = seq(0, 1000, by = 100), right = FALSE, include.lowest = TRUE)),
  levels = unique(cut(seq(0, 1000, by = 1), breaks = seq(0, 1000, by = 100), right = FALSE, include.lowest = TRUE))
))

# Object that will divide the genes in groups depending on the number of DEAs they have appeared as a DEG
gene_count_summary <- gene_count_df_fdr %>%
  mutate(Count_Bin = cut(Count, breaks = seq(0, 1000, by = 100), right = FALSE, include.lowest = TRUE)) %>%
  group_by(Count_Bin) %>%
  summarize(Num_Genes = n()) %>%
  right_join(bins, by = "Count_Bin") %>%
  replace_na(list(Num_Genes = 0))

# Plot the distribution
ggplot(gene_count_summary, aes(x = Count_Bin, y = Num_Genes)) +
  geom_bar(stat = "identity", fill = "steelblue") +
  labs(
    title = "Distribution of Gene Counts by Range",
    x = "Count Range",
    y = "Number of Genes"
  ) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

The object `gene_count_df_fdr` only says how many times each gene has appeared as differentially expressed, whether they are overexpressed or underexpressed in tumor samples with respect to healthy samples is still unknown. For this, the object `results_fdr` will be used, as it holds the fold-change of each gene in each DEA. Whether the fold-change is positive or negative will tell if the gene is overexpressed or underexpressed, respectively.  
```{r expression}
# Object where the number of times each gene appears as overexpressed or underexpressed will be stored
expression_matrix<-matrix(ncol = 2, nrow = 0)

# Loop obtaining the values
for (i in 1:nrow(gene_count_df_fdr)){
  overexpressed<- nrow(results_fdr %>% filter(symbol==gene_count_df_fdr[i,"Gene"] & logFC>0))
  underexpressed<- nrow(results_fdr %>% filter(symbol==gene_count_df_fdr[i,"Gene"] & logFC<0))
  temp_matrix<-matrix(c(overexpressed, underexpressed), ncol=2, nrow=1)
  expression_matrix<- rbind(expression_matrix, temp_matrix)
}

# Adding the values to the count table
gene_count_df_fdr<-cbind(gene_count_df_fdr,expression_matrix)
colnames(gene_count_df_fdr) <- c("Gene", "Count", "overexpressed", "underexpressed")
```

The percentage of the times each gene has come up as a DEG will be obtained and added to the count table and plotted.
```{r percentage}
# Obtaining the percentages
gene_count_df_fdr <- gene_count_df_fdr %>% mutate(percentage=((Count/1000)*100))

# Plot the percentages
ggplot(gene_count_df_fdr, aes(x = Gene, y = percentage)) +
  geom_bar(stat = "identity", position = "identity") +
  scale_fill_manual(values = c("black")) +
  labs(y = "Percantage of positive analyses", x = "Gene") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 90, hjust = 1))
```

To know if all the times they appear as DEG, the change has the same direction, alas, to see if in all cases, the results give underexpression or overexpression, a plot will be created with the count for overexpression and underexpression for each gene. 
```{r image_2}
# Select the important variables and modify the table to ease the plotting and the visualizing 
data_long <- gene_count_df_fdr %>%
  select(Gene, overexpressed, underexpressed) %>%  
  pivot_longer(cols = c(overexpressed, underexpressed), 
               names_to = "ExpressionType", 
               values_to = "Count") %>%
  mutate(Count= ifelse(ExpressionType == "underexpressed", -Count, Count))

# Create the side-by-side bar plot
ggplot(data_long, aes(x = Gene, y = Count, fill = ExpressionType)) +
  geom_bar(stat = "identity", position = position_dodge(width = 0.8)) +
  scale_fill_manual(values = c("overexpressed" = "blue", "underexpressed" = "red")) +
  labs(x = "Gene", y = "Count", title = "Overexpressed vs Underexpressed Counts for Each Gene") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

The next step is to see the magnitude of the changes in expression for the target genes.
```{r average_logFC}
# Object where the average logFC changes
avg_logFC <- matrix(ncol=1, nrow=0)
colnames(avg_logFC)<- c("average_logFC")

# Obtaining the average logFC 
for (i in 1: nrow(gene_count_df_fdr)){
  avg_logFC<-rbind(avg_logFC, mean((results%>% filter(symbol==gene_count_df[i, "Gene"]))$logFC))
}

# Adding the values to the count matrix
gene_count_df_fdr<-cbind(gene_count_df_fdr, avg_logFC)

# Plotting the average logFC
ggplot(gene_count_df_fdr, aes(x = Gene, y = average_logFC, fill = average_logFC > 0)) +
  geom_bar(stat = "identity", position = "identity") +
  scale_fill_manual(values = c("TRUE"= "blue", "FALSE"="red")) +
  labs(y = "average logFC", x = "Gene") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 90, hjust = 1))
```

# Exporting the results

Finally, the count table with the additional information that have been calculated is exported:

```{r export_data}
write.xlsx(gene_count_df_fdr, "Colon_DEA_group.xlsx")
```




