---
title: "Colon_DEA"
author: "Narine Fischer"
date: "2025-09-01"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Introduction

The following workflow uses R language [@rsoftware] to carry out a Differential Expression Analysis (DEA) using data obtained from the Genomic Data Commons Data Portal [@gdcportal]. RNA-seq expression data in samples of patients with Colon Cancer (CC) was analyzed against healthy samples.

## Needed libraries

The libraries used come from both CRAN [@cran] and Bioconductor [@bioconductor]. 

```{r libraries}
library(TCGAbiolinks)
library(edgeR)
library(geneplotter)
library(limma)
library(sva)
library(dplyr)
library(ggplot2)
library(tidyr)
library(rlist)
library(openxlsx)
library(SummarizedExperiment)
library(AnnotationDbi)
library(org.Hs.eg.db)
```

## Download data

## Data download

As previously stated, the data will be obtained from the Genomic Data Commons data portal [@gdcportal], a repository with data from projects studying cancer. In this case, the data will be downloaded from the project TCGA-COAD, with data specific to Colon cancer. Therefore, The results published here are in whole or part based upon data generated by The Cancer Genome Atlas managed by the NCI and NHGRI. Information about TCGA can be found at http://cancergenome.nih.gov .The RNA-seq data obtained during this project will be the one selected. For this step, the package `TCGAbiolinks` will be used [@TCGAbiolinks].
```{r download_data }
# Generate query to obtain the RNA-Seq data from the TCGA-GBM project
query_rnaseq <- GDCquery(
  project = "TCGA-COAD",
  data.category = "Transcriptome Profiling",
  data.type = "Gene Expression Quantification",
  experimental.strategy = "RNA-Seq",
  access = "open"
)

# Download the RNA-Seq data
GDCdownload(query_rnaseq, method = "client")

# Prepare RNA-Seq data for analysis, creating a SummarizedExperiment object
rnaseq_data <- GDCprepare(query_rnaseq)
```

# Quality assessment and processing of data

## Observation of possible confounding variables

As the intent of this workflow is to compare primary tumor samples against healthy samples, the `rnaseq_data` will be filtered to only obtain these two types of samples. Moreover a DGELISt object will be created for further analyses

```{r filter_rnsaseq_data}
# Filter for healthy and primary tumor samples
se<- rnaseq_data[, colData(rnaseq_data)$definition %in% c("Solid Tissue Normal", "Primary solid Tumor")]

# Create a DGEList object for the analysis
dge <- DGEList(counts=assay(se), genes=rowData(se),
                 samples=colData(se))
```

The sequencing depth per sample will be mapped and colored according to a variable that is believed that might cause a confounding effect. The first variable to be observed is `gender`. The aim is to see if there is any difference in sequencing depth due to gender, this should be done with any variable that wants to be investigated. A graph with the ordered sequencing depths and colored by `gender` is created.
```{r quality_assessment_gender}
# Quality assessment: depth sequencing - gender
ord <- order(dge$sample$lib.size)
se$gender<-factor(se$gender)
barplot(dge$sample$lib.size[ord]/1e6, las=1, ylab="Millions of reads",
        xlab="Samples", col=c("red", "blue", "white")[se$gender[ord]])
legend("topleft", c("female", "male", "not reported"), fill=c("red", "blue", "white"), inset=0.01)
```
There are some considerable differences in sequencing depth per sample, ranging from less than 20 to more than 80 million reads; therefore, normalization within samples is needed. In addition, it doesn't appear to be differences in sequencing depth due to gender.

The next variable to be assessed is `race`.
```{r quality_assessment_race}
# Quality assessment: depth sequencing - race. A graph with the ordered sequencing depths and colored by race is created.
ord <- order(dge$sample$lib.size)
se$race<-factor(se$race)
barplot(dge$sample$lib.size[ord]/1e6, las=1, ylab="Millions of reads",
        xlab="Samples", col=c("green", "yellow", "#ED3EF7","white", "#FF885B")[se$race[ord]])
legend("topleft", c("american indian or alaska native","asian", "black or african american", "not reported", "white"), fill=c("green", "yellow", "#ED3EF7", "white", "#FF885B"), inset=0.01)
```

Whereas gender was more homogeneous, there are clear differences between races. "Not reported" samples are mainly in the lower side of sequencing depth, while the rest of samples accumulate on the top of the sequencing depth. Race will be taken as a covariate.

It will also be seen if there is any distinction in sequencing depth depending on the target variable, `definition`.
```{r quality_assessment_definition}
# Quality assessment: depth sequencing - definition. A graph with the ordered sequencing depths and colored by definition is created
se$definition<-factor(se$definition)
barplot(dge$sample$lib.size[ord]/1e6, las=1, ylab="Millions of reads",
        xlab="Samples", col=c("brown", "lightpink")[se$definition[ord]])
legend("topleft", c("Primary solid Tumor", "Solid Tissue Normal"), fill=c("brown", "lightpink"), inset=0.01)
```

It can be seen that Solid tissue normal is mainly present in the higher region of sequencing depth, confirming that normalization is necessary.

## Normalization I: Within-sample

In order to adjust for the differences in sequencing depth, normalization within samples will be carried out. For this, counts per million (CPM) will be calculated. Moreover, the logarithm with base 2 will be carried out to reduce variability and make the data more suitable for statistical analysis.

```{r within_sample_normalization}
# Calculate the log2 of the CPM and add it to the SummarizedExperiment object 
assays(se)$logCPM <- cpm(dge, log=TRUE, prior.count=0.25)
```

The distribution of the expression of the `logCPM` values is plotted to identify if there are any samples that deviates from the rest.

```{r QA_sample_distribution}
# Representation of the distribution of logCPM
par(mfrow=c(1, 2), mar=c(4, 5, 4, 4))
multidensity(as.list(as.data.frame(assays(se)$logCPM)),
               xlab="log2 CPM", legend=NULL,
               main="", cex.axis=1.2, cex.lab=1.5, las=1)
boxplot(assays(se)$logCPM, col="gray", xlab="Samples",
               ylab=expression(log[2] * "CPM"), cex.axis=1.2, cex.lab=1.5, las=1)
```

It appears that there are some unsuitable samples . These samples will be removed.

```{r detecting_samples_to_remove}
num_matrix <- as.matrix(assay(se, "logCPM"))

density_0<- function(x){
  d <-density(x, n=4096)
  # infer density at -1
  unname(approx(d$x, d$y, xout=-1, rule=2)$y)
}

densities<- vapply(seq_len(ncol(num_matrix)), function(j) density_0(num_matrix[,j]), numeric(1))

names(densities) <- colnames(num_matrix)

samples_to_erase <- names(densities[densities >0.05])

# replot to mkae sure the correct samples have been identified

replot_density <- function(full_matrix, highlight){
  # add smoothness to curves
  style<-par(lend=2)
  on.exit(par(style))
  
  # create a blank graph 
  plot (0,0,type="n", xlab="log2 CPM", ylab="Density", xlim=range(full_matrix), ylim=c(0,0.5))
  
  #fill the graph
  for (s in colnames(full_matrix)){
    densities<-density(full_matrix[,s], n=2048)
    #choose colour
    colour<- if (s %in% highlight) "red" else "grey"
    lines(densities$x, densities$y, lwd=if(s %in% highlight) 2.5 else 1, col=colour)
  }
  legend("topright", lwd=c(2.5,1), col=c("red", "grey"), legend= c("eliminate", "save"), bty="n")
}

replot_density(num_matrix, samples_to_erase)
```

It can be seen that the samples detected indeed are the samples to remove, therefore, the removal is carried out. In addition, the samples were searched in the sample datatable to see if they were cancer samples, or healthy samples. All of them are cancer samples, making the total of samples after removal: 79 cancer samples and 197 healthy samples.

```{r remove_bad_samples}
se <- se[, !colnames(se) %in% samples_to_erase]
dge <- DGEList(counts=assay(se), genes=rowData(se),
                 samples=colData(se))
```

Finally, we replot the multidensity plot to make sure no other unsuitable samples remain.
```{r make_sure}
# Representation of the distribution of logCPM
par(mfrow=c(1, 2), mar=c(4, 5, 4, 4))
multidensity(as.list(as.data.frame(assays(se)$logCPM)),
               xlab="log2 CPM", legend=NULL,
               main="", cex.axis=1.2, cex.lab=1.5, las=1)
boxplot(assays(se)$logCPM, col="gray", xlab="Samples",
               ylab=expression(log[2] * "CPM"), cex.axis=1.2, cex.lab=1.5, las=1)
```

It can be seen that there are still some unsuitable samples. A second round of elimination is carried out

```{r detecting_samples_to_remove}
num_matrix <- as.matrix(assay(se, "logCPM"))

density_0<- function(x){
  d <-density(x, n=4096)
  # infer density at -3
  unname(approx(d$x, d$y, xout=-3, rule=2)$y)
}

densities<- vapply(seq_len(ncol(num_matrix)), function(j) density_0(num_matrix[,j]), numeric(1))

names(densities) <- colnames(num_matrix)

samples_to_erase <- names(sort(densities, decreasing=FALSE))[1:3]

# replot to mkae sure the correct samples have been identified

replot_density <- function(full_matrix, highlight){
  # add smoothness to curves
  style<-par(lend=2)
  on.exit(par(style))
  
  # create a blank graph 
  plot (0,0,type="n", xlab="log2 CPM", ylab="Density", xlim=range(full_matrix), ylim=c(0,0.5))
  
  #fill the graph
  for (s in colnames(full_matrix)){
    densities<-density(full_matrix[,s], n=2048)
    #choose colour
    colour<- if (s %in% highlight) "red" else "grey"
    lines(densities$x, densities$y, lwd=if(s %in% highlight) 2.5 else 1, col=colour)
  }
  legend("topright", lwd=c(2.5,1), col=c("red", "grey"), legend= c("eliminate", "save"), bty="n")
}

replot_density(num_matrix, samples_to_erase)
```

```{r remove_bad_samples}
se <- se[, !colnames(se) %in% samples_to_erase]
dge <- DGEList(counts=assay(se), genes=rowData(se),
                 samples=colData(se))
```

```{r detecting_samples_to_remove}
num_matrix <- as.matrix(assay(se, "logCPM"))

density_0<- function(x){
  d <-density(x, n=4096)
  # infer density at -1
  unname(approx(d$x, d$y, xout=-2.5, rule=2)$y)
}

densities<- vapply(seq_len(ncol(num_matrix)), function(j) density_0(num_matrix[,j]), numeric(1))

names(densities) <- colnames(num_matrix)

samples_to_erase <- names(densities[densities >0.06])

# replot to mkae sure the correct samples have been identified

replot_density <- function(full_matrix, highlight){
  # add smoothness to curves
  style<-par(lend=2)
  on.exit(par(style))
  
  # create a blank graph 
  plot (0,0,type="n", xlab="log2 CPM", ylab="Density", xlim=range(full_matrix), ylim=c(0,0.5))
  
  #fill the graph
  for (s in colnames(full_matrix)){
    densities<-density(full_matrix[,s], n=2048)
    #choose colour
    colour<- if (s %in% highlight) "red" else "grey"
    lines(densities$x, densities$y, lwd=if(s %in% highlight) 2.5 else 1, col=colour)
  }
  legend("topright", lwd=c(2.5,1), col=c("red", "grey"), legend= c("eliminate", "save"), bty="n")
}

replot_density(num_matrix, samples_to_erase)
```

```{r remove_bad_samples}
se <- se[, !colnames(se) %in% samples_to_erase]
dge <- DGEList(counts=assay(se), genes=rowData(se),
                 samples=colData(se))
```

```{r make_sure}
# Representation of the distribution of logCPM
par(mfrow=c(1, 2), mar=c(4, 5, 4, 4))
multidensity(as.list(as.data.frame(assays(se)$logCPM)),
               xlab="log2 CPM", legend=NULL,
               main="", cex.axis=1.2, cex.lab=1.5, las=1)
boxplot(assays(se)$logCPM, col="gray", xlab="Samples",
               ylab=expression(log[2] * "CPM"), cex.axis=1.2, cex.lab=1.5, las=1)
```

Although it is much better, the last two samples that are deemed unsuitable will be eliminated.

```{r detecting_samples_to_remove}
num_matrix <- as.matrix(assay(se, "logCPM"))

density_0<- function(x){
  d <-density(x, n=4096)
  # infer density at -3
  unname(approx(d$x, d$y, xout=-2.5, rule=2)$y)
}

densities<- vapply(seq_len(ncol(num_matrix)), function(j) density_0(num_matrix[,j]), numeric(1))

names(densities) <- colnames(num_matrix)

samples_to_erase <- names(sort(densities, decreasing=FALSE))[1:2]

# replot to mkae sure the correct samples have been identified

replot_density <- function(full_matrix, highlight){
  # add smoothness to curves
  style<-par(lend=2)
  on.exit(par(style))
  
  # create a blank graph 
  plot (0,0,type="n", xlab="log2 CPM", ylab="Density", xlim=range(full_matrix), ylim=c(0,0.5))
  
  #fill the graph
  for (s in colnames(full_matrix)){
    densities<-density(full_matrix[,s], n=2048)
    #choose colour
    colour<- if (s %in% highlight) "red" else "grey"
    lines(densities$x, densities$y, lwd=if(s %in% highlight) 2.5 else 1, col=colour)
  }
  legend("topright", lwd=c(2.5,1), col=c("red", "grey"), legend= c("eliminate", "save"), bty="n")
}

replot_density(num_matrix, samples_to_erase)
```

```{r remove_bad_samples}
se <- se[, !colnames(se) %in% samples_to_erase]
dge <- DGEList(counts=assay(se), genes=rowData(se),
                 samples=colData(se))
```

```{r make_sure}
# Representation of the distribution of logCPM
par(mfrow=c(1, 2), mar=c(4, 5, 4, 4))
multidensity(as.list(as.data.frame(assays(se)$logCPM)),
               xlab="log2 CPM", legend=NULL,
               main="", cex.axis=1.2, cex.lab=1.5, las=1)
boxplot(assays(se)$logCPM, col="gray", xlab="Samples",
               ylab=expression(log[2] * "CPM"), cex.axis=1.2, cex.lab=1.5, las=1)
```

It can be confirmed that no more unsuitable samples remain. Therefore, the next step can begin. In the end, the samples saved are:

```{r table_se_definition}
table(factor(se$definition))
```

Before, there were 481 tumor samples, and 41 healthy samples. Now, 452 tumor samples remain, and 41 healthy samples remain. Therefore, all samples erased were tumor samples. 

## Filtering of lowly-expressed genes

The next step will be filtering out the lowly-expressed genes, as they can introduce artifacts in the analysis. First, a representation of the gene expression is done.

```{r QA_gene_distribution}
# Calculate the average level of expression of each gene
avgexp <- rowMeans(assays(se)$logCPM)

# Plot the average levels of expression 
hist(avgexp, xlab=expression(log[2] * "CPM"), main="", las=1, col="gray")
abline(v=0, col="red", lwd=2)
```

In this case, the cutoff has been calculated according to Chen et al. (2016), 10 reads per million in the sample with the lowest depth. Genes kept must have this minimum level of expression in at least as many samples as the smallest group of comparison (healthy samples, 41 samples)

```{r LE_genes_removal}
# Create mask using calculated minimum log2CPM as cutoff
cpmcutoff <- round(10/min(dge$samples$lib.size/1e6),digits=1)
nsamplescutoff <- min(table(se$definition))

mask <- rowSums(cpm(dge) > cpmcutoff) >= nsamplescutoff

# Create new objects with the filtered genes 
se.filt <- se[mask, ]
dge.filt <- dge[mask, ]
dim(se.filt)
```

A total of 16988 genes that meet the cutoff are kept.

Plotting again the gene expression plot, coloring the 16988 genes in red, the correct selection is confirmed
```{r filtered_genes_representation}
par(mar=c(4, 5, 1, 1))
h <- hist(avgexp, xlab=expression("Expression level (" * log[2] * "CPM)"),
          main="", las=1, col="grey", cex.axis=1.2, cex.lab=1.5)
x <- cut(rowMeans(assays(se.filt)$logCPM), breaks=h$breaks)
lines(h$mids, table(x), type="h", lwd=10, lend=1, col="red")
legend("topright", c("All genes", "Filtered genes"), fill=c("grey", "red"))
```

## Normalization II: between-sample

The next step will be normalizing between samples. The TMM method will be used.

```{r normalization_bet_sample}
dge.filt <- calcNormFactors(dge.filt, method = 'TMM')
```

The raw data is substituted by the normalized data

```{r raw_to_normalized}
# Replace the raw log2 CPM units in the corresponding assay element of the `SummarizedExperiment` object, by the normalized ones.
assays(se.filt)$logCPM <- cpm(dge.filt, log=TRUE,
                              normalized.lib.sizes=TRUE)
```

A boxplot is created to see if there are any significant differences in expression between samples to see if there is any outlier between them

```{r repeat_boxplot}
boxplot(assays(se.filt)$logCPM, col="gray", xlab="Samples",
        ylab=expression(log[2] * "CPM"), las=1)
```

## QA: MA-plots

To compare, first the individual MA-plots prior to filtering will be plotted. 
```{r MA_plot_before_filter}
# Print them 4 at a time
par(mfrow=c(2, 2), mar=c(4, 5, 1, 1))

# MA-plot generation
for (i in 1:ncol(se)) {
  A <- rowMeans(assays(se)$logCPM)
  M <- assays(se)$logCPM[, i] - A
  smoothScatter(A, M, main=colnames(se)[i], las=1)
  abline(h=0, col="blue", lwd=2)
  lo <- lowess(M ~ A)
  lines(lo$x, lo$y, col="red", lwd=2)
}
```

Then, the individual MA-plots after filtering and normalization against the average values will be plotted 

```{r MA_plot_after_filter}
# Print 4 graphs at a time
par(mfrow=c(2, 2), mar=c(4, 5, 1, 1))

# MA-plot generation
for (i in 1:ncol(se.filt)) {
  A <- rowMeans(assays(se.filt)$logCPM)
  M <- assays(se.filt)$logCPM[, i] - A
  smoothScatter(A, M, main=colnames(se.filt)[i], las=1)
  abline(h=0, col="blue", lwd=2)
  lo <- lowess(M ~ A)
  lines(lo$x, lo$y, col="red", lwd=2)
}
```

Due to the difference in number of samples, it's difficult to see differences in some samples.

Next, a smear plot will be done to compare `Primary solid Tumor` and `Solid Tissue Normal`as groups. As before, a graph before filtering and another after filtering will be carried out.
```{r comparing_by_groups}
# Print the graphs together
par(mfrow=c(2, 1), mar=c(4, 5, 1, 1))

# Assign the type of samples to the DGEList objects
se$definition <- relevel(factor(se$definition), ref="Solid Tissue Normal")
se.filt$definition <- relevel(factor(se.filt$definition), ref="Solid Tissue Normal")
dge$samples$group <- se$definition
dge.filt$samples$group <- se$definition

# Smear plot before filtering
plotSmear(dge, lowess=TRUE, las=1, cex.lab=1.5, cex.axis=1.2)
abline(h=0, col="blue", lwd=2)

# Smear plot after filtering
plotSmear(dge.filt, lowess=TRUE, las=1, cex.lab=1.5, cex.axis=1.2)
abline(h=0, col="blue", lwd=2)

```

It can be seen that in the lower expression, there is an overexpression in Differentially expressed genes, however, this image could be due to the difference in number of samples.

## Batch identification and experimental design

In order to detect possible batches that can interfere in the analysis, a hierarchical  clustering is carried out.

```{r batch_identification_hierarchical clustering,  fig.height=5, fig.width=8, dpi=100,}
# Calculate log CPM values with a prior.count of 3 to moderate extreme fold-changes produced by low counts
logCPM <- cpm(dge.filt, log=TRUE, prior.count=3)

# Calculate the distance matrix using Spearman correlation
d <- as.dist(1-cor(logCPM, method="spearman"))

# Perform hierarchical clustering
sampleClustering <- hclust(d)

# To observe gender as a possible batch effect, an object will be created for later use to color each sample's gender
batch <- as.integer(se.filt$race)
names(batch) <- colnames(se.filt)

# Create a dendrogram object with the hierarchical clustering 
sampleDendrogram <- as.dendrogram(sampleClustering, hang=0.1)

# Create the labels that will be seen in the dendrogram
outcome <- se.filt$shortLetterCode
names(outcome) <- colnames(se.filt)

# Modify dendrogram labels and color
sampleDendrogram <- dendrapply(sampleDendrogram,
                               function(x, batch, labels) {
                                 if (is.leaf(x)) {
                                   attr(x, "nodePar") <- list(lab.col=as.vector(batch[attr(x, "label")]))
                                   attr(x, "label") <- as.vector(labels[attr(x, "label")])
                                 }
                                 x
                               }, batch, outcome)

# Plot the dendrogram
plot(sampleDendrogram, main="Hierarchical clustering of samples",
     cex=0.7)
legend("topright", levels(se.filt$race),
       fill=seq_len(nlevels(se.filt$race)))
```

It can be seen that healthy samples are clusterted together, and while some tumor samples are more closely related to them, they are still separated phylogenetically
For race, it doesn't seem to be any clustering, however, as "american indian or alaska native" only appear in the healthy samples, it will still be taken into account as a confounding effect.

Now we colour by gender.

```{r batch_identification_hierarchical clustering,  fig.height=5, fig.width=8, dpi=100,}
# Calculate log CPM values with a prior.count of 3 to moderate extreme fold-changes produced by low counts
logCPM <- cpm(dge.filt, log=TRUE, prior.count=3)

# Calculate the distance matrix using Spearman correlation
d <- as.dist(1-cor(logCPM, method="spearman"))

# Perform hierarchical clustering
sampleClustering <- hclust(d)

# To observe gender as a possible batch effect, an object will be created for later use to color each sample's gender
batch <- as.integer(se.filt$gender)
names(batch) <- colnames(se.filt)

# Create a dendrogram object with the hierarchical clustering 
sampleDendrogram <- as.dendrogram(sampleClustering, hang=0.1)

# Create the labels that will be seen in the dendrogram
outcome <- se.filt$shortLetterCode
names(outcome) <- colnames(se.filt)

# Modify dendrogram labels and color
sampleDendrogram <- dendrapply(sampleDendrogram,
                               function(x, batch, labels) {
                                 if (is.leaf(x)) {
                                   attr(x, "nodePar") <- list(lab.col=as.vector(batch[attr(x, "label")]))
                                   attr(x, "label") <- as.vector(labels[attr(x, "label")])
                                 }
                                 x
                               }, batch, outcome)

# Plot the dendrogram
plot(sampleDendrogram, main="Hierarchical clustering of samples",
     cex=0.7)
legend("topright", levels(se.filt$gender),
       fill=seq_len(nlevels(se.filt$gender)))
```

It can be seen that gender is not clustered.

Another way to determine batches is through a Multidimensional Scaling (MDS) plot.

```{r batch_identificatio_nmps_graph}

batch <- as.integer(se.filt$race)
names(batch) <- colnames(se.filt)

# Creating the labels of the samples
outcome <- se.filt$shortLetterCode
names(outcome) <- colnames(se.filt)

# Creating the MDS plot
plotMDS(dge.filt, labels=outcome, col=batch)
legend("bottomright", levels(se.filt$race),
       fill=seq_len(nlevels(se.filt$race)), inset=0.05)
```

Now, we plot with gender

```{r batch_identificatio_nmps_graph}

batch <- as.integer(se.filt$gender)
names(batch) <- colnames(se.filt)

# Creating the labels of the samples
outcome <- se.filt$shortLetterCode
names(outcome) <- colnames(se.filt)

# Creating the MDS plot
plotMDS(dge.filt, labels=outcome, col=batch)
legend("bottomright", levels(se.filt$gender),
       fill=seq_len(nlevels(se.filt$gender)), inset=0.05)
```

It can be seen a difference between tumor samples and healthy samples, as they cluster separately. In addition, gender does not seem to cluster.

## Differential expression analysis

In order to start the DEA; in previous steps, the levels of the target variable were modified to make the healthy samples as references, this way, all results will be of the tumor samples compared to the healthy samples. However, teh releveling will be done again to ensure that there are no mistakes. Moreover, the model for the analysis is created, as stated before, race will be taken into account.

```{r define_the_reference}
# Define the reference
se.filt$definition <- relevel(se.filt$definition, ref="Solid Tissue Normal")
dge.filt$samples$group <- relevel(dge.filt$samples$group, ref="Solid Tissue Normal")
# Creating the model for the analysis
mod <- model.matrix(~ definition + race, data=colData(se.filt))
head(mod, 15)
```

Although some possible confounding effects were tested for, there might be others that are being missed, hidden batch effects. To avoid this, a Surrogate Variable Analysis (SVA) will be carried out.

```{r SVA}
# Creating a model to compare the previously created against
mod0 <- model.matrix(~ race, colData(se.filt))

# Carry out the SVA analysis
sv <- sva(assays(se.filt)$logCPM, mod=mod, mod0=mod0)
```

Thanks to the results fo the analysis, 50 significant surrogate variables are obtained.

In order to see their significance, a Singular Value Decomposition (SVD) is done to see the amount of contribution to the variance by each surrogate variable

```{r SVA_SVD}
# Perform SVD on the logCPM values
svd_result <- svd(scale(assays(se.filt)$logCPM))
eigenvalues <- svd_result$d^2

# Plot eigenvalues
plot(eigenvalues, type = "b", xlab = "Component", ylab = "Eigenvalue", main = "SVD Eigenvalues")

```

Despite sva telling it found 50 significant surrogate variables, the first five surrogate variables accounts for much of the variance unrelated to the "Tumor vs Healthy" differences. Although seeing this results the 50 surrogate variable swill be added to the model to better help mitigate the noise and potential confounding effects.

The model is modified to take the surrogate variables into account. Then the DEA is carried out with `voomLmFit`from the package `EdgeR` [@edgeR], and with `eBayes`from the package `limma` [@limma]

```{r t-statistic}
# Addition of the SVA surregate variables
mod <- cbind(mod, sv$sv)

# Carry out the Differential Expression Analysis
fit <- voomLmFit(dge.filt, mod, block= se$patient)
fit <- eBayes(fit)
```

The results of the DEA can be found in the object `fit`.

In `fit`, the genes do not appear as their symbols, to make it more understandable, the Ensembl code [@Ensembl2024] will be changed by the genes' symbols. 

```{r gene_data}
# 1) Get Ensembl gene IDs (strip version if present)
ens <- rowData(se.filt)$gene_id
ens <- sub("\\.\\d+$", "", ens)  # drop ".12" version suffixes

# 2) Map to SYMBOL, ENTREZID, chromosome, and a readable name
symbol <- mapIds(org.Hs.eg.db, keys = ens, keytype = "ENSEMBL", column = "SYMBOL",   multiVals = "first")
entrez <- mapIds(org.Hs.eg.db, keys = ens, keytype = "ENSEMBL", column = "ENTREZID", multiVals = "first")
chr    <- mapIds(org.Hs.eg.db, keys = ens, keytype = "ENSEMBL", column = "CHR",      multiVals = "first")
gname  <- mapIds(org.Hs.eg.db, keys = ens, keytype = "ENSEMBL", column = "GENENAME", multiVals = "first")

# 3) Build the data.frame for limma
genesmd <- data.frame(
  chr         = unname(chr),
  symbol      = unname(symbol),                     # or rowData(se.filt)$Gene_name if you prefer
  description = unname(gname),                      # a readable gene name/description
  EntrezID    = unname(entrez),
  stringsAsFactors = FALSE,
  row.names = rownames(se.filt)
)

# 4) Attach to the limma fit
fit$genes <- genesmd
```


Before proceeding, correction for multiple testing is necessary, for it, the function `topTable()` is used. This function uses Benjamini-Hochberg FDR (Flase Discovery Rate) as method for correction.

Moreover, the amount of genes that are overexpressed, underexpressed or stays the same in healthy samples (intercept) will also be printed.

```{r multiple_test_correction}
# Multiple testing correction
tt <- topTable(fit, coef=2, n=Inf)

# Genes divided by change in expression
res <- decideTests(fit, p.value=0.1)
print(summary(res))
```

A plot with the distribution of the P.values for each gene is created.

```{r pvalues}
# Create a histogram with the P.values
h <- hist(tt$P.Value, xlab="Raw P-values", main="",
          las=1)
```

It can be seen that many genes have low p-values, signaling to differences in expression.

A volcano plot is also created to see the distribution of the differentially expressed genes, either underexpressed or overexpressed.

```{r volcano}
# Create the volcano plot
volcanoplot(fit, coef=2, highlight=10, names=fit$genes$symbol,
            main="", las = 1)
```

In addition, an MA-plot will also be created to also represent the differentially expressed genes distribution.
```{r MA_plots_after}
# Obtaining the name of the top 10 DEG
top10 <- order(fit$lods[, 2], decreasing=TRUE)[1:10]

# Filter for genes with less than a 0.1 adjusted P.value
DEgenes <- rownames(tt)[tt$adj.P.Val < 0.1]

# Createing the MA-plot
limma::plotMA(fit, coef=2, status=rownames(fit$lods) %in% DEgenes, legend=FALSE,
              main="", hl.pch=46, hl.cex=4, bg.pch=46, bg.cex=3, las =1)
text(fit$Amean[top10], fit$coef[top10, 2], fit$genes$symbol[top10], cex=0.5, pos=4)
```


To have the complete list of differentially expressed genes, a final filter is undergone, only those that after the correction have an adjusted P.value of less than 0.1 will be kept.
```{r DGE_list}
DEGtable <- tt %>% filter(adj.P.Val < 0.1) %>% arrange(P.Value)
rownames(DEGtable) <- 1:nrow(DEGtable)
```

# Exporting the results

Finally, the count table with the additional information that have been calculated is exported:
```{r export_data}
write.xlsx(DEGtable, "Colon_DEGtable_primary_healthy.xlsx")
```
